---
title: "Description of Capstone Package"
author: "Federico Sensi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The earthquaketools R package provides a range of functions that operate on (clean), and visualize,
earthquake data from the U.S. National Oceanographic and Atmospheric Administration (NOAA) database: 
https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1

This vignette provides an overview of each of the functions provided, with examples.

## eq_clean_data() Function

 This function takes the raw data frame of NOAA earthquake data and cleans it.
 Data (tab delimited) is available for download at:
 \href{https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1}{Significant
 EarthQuake Database}
 The clean data frame have the following:
 1. A date column created by uniting the year, month, day and converting it to
    the Date class
 2. LATITUDE and LONGITUDE columns converted to numeric class

```{r}
library(capstoneproject)
# Read the raw data from package
data(earthquakes)
# Clean it up for use by the visualization functions
quakes_clean <- eq_clean_data(earthquakes)
```

## eq_clean_location() Function
This function cleans the LOCATION_NAME observation in a NOAA significant Earthquakes data set.

```{r}
library(capstoneproject)
# Read the raw data from package
data(earthquakes)
# Clean LOCATION_NAME for use by the visualization functions
quakes_loc_clean <- eq_clean_location(earthquakes)
```


## Functions to Visualize the Earthquake Data

Four functions have been created that help visualize the data:

Function Name | Description
--------------|------------
geom_timeline() | Adds an earthquake timeline plot to a ggplot2/grid graphics object
geom_timeline_label() | Label Largest Earthquakes on Timeline created by geom_timeline()
eq_map() | Creates an interactive leaflet map showing the location of earthquakes in the given quakes data set.
eq_create_label() | Creates a more descriptive and HTML-formatted popup label to be used in eq_map function

## geom_timeline() Function

geom_timeline shows a timeline of NOAA Significant earthquakes, plotting individual 
countries along the y-axis and dates along the x-axis. The size of the points is relative to 
the earthquakes magnitude, and the color is related to the total number of deaths.

```{r, eval = TRUE, fig.width=7}
library(capstoneproject)
library(dplyr)
library(ggplot2)

data(earthquakes)
quakes_clean <- eq_clean_data(earthquakes)

 quakes_clean %>%
    filter(COUNTRY %in% c('ITALY', 'GREECE') & DATE > '1990-01-01' & !is.na(EQ_PRIMARY)) %>%
    ggplot() +
    geom_timeline(aes(x = DATE, y = COUNTRY, color = TOTAL_DEATHS,
                      size = EQ_PRIMARY)) +
    scale_size_continuous(name = 'Richter scale value') +
    scale_color_continuous(name = 'Nbr of Deaths')
```

## geom_timeline_label() Function

Function geom_timeline shows a timeline of NOAA Significant earthquakes,
plotting individual countries along the y-axis and dates along the x-axis.
The size of the points is relatative to the earthquakes magnitude, and
the color is related to the total number of deaths.

```{r, eval = TRUE, fig.width=7}
library(capstoneproject)
library(dplyr)
library(ggplot2)
library(grid)
 
 data(earthquakes)
 quakes_clean <- eq_clean_data(earthquakes)

  quakes_clean %>%
     dplyr::filter(COUNTRY %in% c('ITALY', 'GREECE') & DATE > '2000-01-01' & !is.na(EQ_PRIMARY)) %>%
     ggplot() +
     geom_timeline(aes(x = DATE, y = COUNTRY, color = TOTAL_DEATHS,
                       size = EQ_PRIMARY)) +
     geom_timeline_label(aes(x = DATE, y = COUNTRY, magnitude = EQ_PRIMARY,
                       label = LOCATION_NAME, n_max = 3)) +
     scale_size_continuous(name = 'Richter scale value') +
     scale_color_continuous(name = 'Nbr of Deaths')
```


## eq_map() Function

This function shows an interactive map of the location of the earthquakes in the given quakes data.  
The size of the circles are proportional to the magnitude of the earthquakes (in the EQ_PRIMARY) variable. 
The map is interactive, and when you click on a link, the popup shows the annotation as specified by the \code{annot_col} variable.



```{r, eval = TRUE, fig.width=7}
library(capstoneproject)
library(lubridate); 
library(dplyr)

data(earthquakes)
quakes_clean <- eq_clean_data(earthquakes)
quakes_it_gr  <-  quakes_clean %>%
  dplyr::filter(COUNTRY %in% c('ITALY', 'GREECE') & DATE > '2000-01-01' & !is.na(EQ_PRIMARY))
eq_map(quakes_it_gr, annot_col = 'DATE')
```

## eq_create_label() Function

If the user desires to show more than just one piece of information about a selected earthquake, the eq_create_label() function may be used. Function eq_create_label creates a more descriptive and HTML-formatted popup label to be used in eq_map.

```{r, eval = TRUE, fig.width=7}
library(capstoneproject)
library(lubridate); 
library(dplyr)

data(earthquakes)
quakes_clean <- eq_clean_data(earthquakes)
quakes_it_gr_popup  <-  quakes_clean %>%
  dplyr::filter(COUNTRY %in% c('ITALY', 'GREECE') & DATE > '2000-01-01' & !is.na(EQ_PRIMARY)) %>%
  dplyr::mutate(popup_text = eq_create_label(.))
eq_map(quakes_it_gr_popup, annot_col = 'popup_text')
```
